\documentclass{jn-pset}
%   \documentclass[solutions]{jn-pset}

\usepackage{ifthen}
\newboolean{versionwithsolutions}
% Uncomment to insert text that should only be there in version WITH solutions
\setboolean{versionwithsolutions}{true}
% Uncomment to insert text that should only be there in version WITHOUT solutions
\setboolean{versionwithsolutions}{false}


%%%%
%%%% QUICK INSTRUCTIONS FOR FORMATTING OF PROBLEMS
%%%%
%    
% Code a problem by
%    \begin{problem}
%    \end{problem}
%
% Code a subproblem inside a problem by (note the percent signs!)
%    \begin{subproblem}%
%        \label{problem:labelhere}%
%        Text here
%    \end{subproblem}
%
% Get a small vertical space by issuing the command
%    \smallskip
%
% For instance, to give a hint to a problem (after having completed
% the problem statement) code in the following way:
%
%    \smallskip
%     \noindent
%     \emph{Hint:}
%     Consider the following super-useful hint for this particular problem...
%

% PROBLEM-SET-SPECIFIC MACROS (UPDATE FOR EACH PROBLEM SET)

\newcommand{\psetno}{3}
\newcommand{\duedate}{Wedneday March 12 at 12:59 CET}

\newcommand{\thresholdforpass}{$120$~points\xspace}

\input{course-macros.tex}

% For getting watermark "DRAFT" across all pages (for instance, 
% when posting preliminary version of problem set)
%    \usepackage{draftwatermark}
%    % \SetWatermarkFontSize{20 cm}
%    \SetWatermarkScale{5}

% For METAPOST logo as \hologo{METAPOST}
%   \usepackage{hologo}

% For TikZ
%   \input{Figures/tikz-packages.tex}

%%%
%%% TITLE
%%%

\author{\courseinstructor}
\course{\coursenamelong{}}
\semester{\courseperiod}
\title{\coursenameshort: Problem Set \psetno}

\begin{document}

\maketitle

\input{IDMA_Pset_GeneralInstructions}


\begin{problem}%
  (40 p)
  Recall that a standard deck of cards has $52$~cards
  partitioned into four \emph{suits}
  (hearts, spades, clubs, and diamonds)
  with $13$~ranks each
  ($2$--$10$ plus jack, queen, king, and ace).
%%
  In this problem, we assume that you are dealt $5$~cards from a
  perfectly shuffled deck  of cards,
  and we wish to analyse the probability of getting flushes and/or straights.
  Note that there are two straights involving the ace: both ace--2--3--4--5 and
  10--jack--queen--king--ace are valid straights.

  \begin{subproblem}
    What is the probability that you get a \emph{flush},
    i.e., 
    $5$~cards of the same suit
    but not all in sequence with respect to rank?
    (Because  five cards of the same suit in sequential rank
    would be  a \emph{straight flush}.)
  \end{subproblem}

\begin{solution}
  The number of flushes, straight or not,
  is obtained by first choosing a suit in $4$~ways,
  and then choosing $5$~cards in $\binom{13}{5}$
  ways.
  After choosing the suit, 
  the number of straight flushes is obtained by
%     choosing the lowest ranked card
%     $2$--$10$ in $9$ possible ways.
  choosing the lowest card in the straight
  $ace$--$10$ in $10$ possible ways.
  Hence, the total number of (non-straight) flushes is
  $
%     4 \bigl( \binom{13}{5} - 9 \bigr)
  4 \bigl( \binom{13}{5} - 10 \bigr)
  $.
  The total possible number of outcomes for a hand of $5$~cards is
  $\binom{52}{5}$.
  Since all outcomes are equally likely, 
  the probability of getting a (non-straight) flush is
  \begin{equation*}
    \frac
%       {4 \bigl( \binom{13}{5} - 9 \bigr)}
    {4 \bigl( \binom{13}{5} - 10 \bigr)}
    {\binom{52}{5}}
    \eqcomma
  \end{equation*}
  i.e., the number of successful outcomes divided by the total number
  of outcomes.
\end{solution}

  \begin{subproblem}
    What is the probability that you get a \emph{straight},
    i.e., 
    $5$~cards of sequential rank
    but not all of the same suit?
    (Because if the latter condition also held, we would again have a
    straight flush.) 
  \end{subproblem}

\begin{solution}
  The number of straights,  flushes or not,
  is obtained by first choosing
%     the rank of the lowest ranked card
%     $2$--$10$ in $9$ possible ways,
  the rank of the lowest card
  $2$--$10$ in $10$ possible ways,
  and then for each of the $5$~cards
  choosing the suit  in a total of  $4^5$~ways.
  To get a straight flush, we can choose the suit of the first card in
  $4$~ways, but then there is no freedom of choice for the rest of the cards.
  Hence, the total number of (non-flush) straights is
  $
%     9 \bigl(4^5 - 4    \bigr)
  10 \bigl(4^5 - 4    \bigr)
  $.
  The total possible number of outcomes for a hand of $5$~cards is
  still
  $\binom{52}{5}$, and so 
  the probability of getting a (non-flush) straight is
  \begin{equation*}
    \frac
%       {  9 \bigl(4^5 - 4    \bigr) }
    {  10 \bigl(4^5 - 4    \bigr) }
    {\binom{52}{5}}
    \eqperiod
  \end{equation*}
\end{solution}


\end{problem}


\begin{problem}%
  (40 p)
%     Show
  Prove mathematically 
  that among all numbers on the form
  $11\ldots{}1
  00\ldots{}0$,
  i.e., numbers consisting of $m$~ones followed by $n$~zeros
  for some $m,n \in  \Nplus$
%     (jumping ahead a bit, later in the course we will introduce
%     the notation $1^m0^n$ to describe
%     the text strings corresponding to such numbers),
  (sometimes notation like $1^m0^n$ is used to describe
  text strings constructed in such a way),
  there is some number that is divisible by~$2025$.
  
  \smallskip
  \noindent
  \emph{Hint:} 
  Look at all numbers $1^m  =  11\ldots{}1$ 
  and consider what their remainders can be modulo~$2025$.

\begin{solution}
    Using the notation introduced within parentheses above,
    among the first $2026$ numbers on the form
    $11\ldots{}1$ 
    there have to be two numbers 
    $1^m$
    and
    $1^n$
    (i.e., with $m$~and $n$~digits~$1$, respectively)
    for $m > n$
    with the same remainder
    modulo~$2025$.
%
    Here we are using the pigonhole principle with the numbers 
    $11\ldots{}1$
    serving
    as pigeons and the possible remainders modulo~$2025$ serving as holes.
%
    But that means that $2025$ divides the difference of these two numbers,
    and subtracting the smaller number from the larger one we get a
    number on the form
    $1^m0^n$
    that is divisible by~$2025$, as desired.

%       
%       As a side note, in a mathematical context the word ``show'' should
%       be understood as a synonym for ``prove''.  If a student ran
%       computer experiments and found a large number on the form $1^m0^n$
%   

    As a side note, if one would run computer experiments to find a
    large number on the form $1^m0^n$
    that the computer claims is divisible by~$2025$, such a number
    cannot just be reported without any further evidence---there is no
    obvious reason why we should believe such a claim. In order to get
    a full score for such a solution,
%       the student also needs
    one would also need
    to state
    clearly the factorization and work out the details to show that it
    is correct. (Another matter is that it should be fairly clear that
    this is not how the problem was intended to be solved, and that
    problems on the exam will not be possible to solve in a similar
    way, so it is arguably a somewhat short-sighted approach to try to
    solve the problem in this way.)
\end{solution}
\end{problem}


\begin{problem}
  (40 p)
  Let $a \in \Rplus$ be any positive real number.
  Show that for any
%     positive 
  integer $n \geq 2$
  there is a rational number
  $\frac{c}{d}$,   %%%   $c/d$,
  $c, d \in \Z$,
  $d \leq n$,
  that approximates~$a$ to within error
  $\frac{1}{dn}$,
  i.e.,
  $
  \Abs{a - \frac{c}{d}} \leq \frac{1}{dn}
  $.

  \smallskip
  \noindent
  \emph{Hint:} 
  Consider the numbers $a, 2a, \ldots, n \cdot a$ 
  and show that one of these numbers is at distance at most $1/n$ from
  some integer.  

%   
%   \abs{da - c} <= 1/n
%   \abs{a - c/d} <= 1/dn
%   
\begin{solution}
  Using the hint, for some $d = 1, \ldots, n$
  we would like to find
  an integer~$c$ such that
  $\abs{d \cdot a - c} \leq \frac{1}{n}$.
  To this end, we will use the pigeonhole principle.

  For every number
  $d \cdot a$, $1 \leq d \leq n$,
  let 
  $c_d = \floor{d \cdot a}$
  be the closest smallest integer.
  Clearly, by construction we have
  $0 \leq d \cdot a - c_d < 1$.


  Divide the interval~$[0,1]$ between $0$ and~$1$
  into $n$~evenly spaced subintervals
  $
  \big[
  \frac{i-1}{n},
  \frac{i}{n}
  \big]
  $
  for $1 \leq i \leq 1$.
  If for some~$d$ and~$c_d$ 
  we have that
  $d \cdot a - c_d$
  is in the subinterval
  $
  \big[
  0,
  \frac{1}{n}
  \big]
  $,
  then fix~$c = c_d$.
  Otherwise, all other subintervals will serve as $n-1$ pigeonholes,
  and the numbers
  $d \cdot a - c_d $
%     $c_d = \floor{d \cdot a}$
  for
  $1 \leq d \leq n$
  will serve as pigeons.
  By the pigeonhole principle,
  there are two numbers
  $c_{d'}$ and~$c_{d''}$
  for
  $d' > d''$
  such that
  $d' \cdot a - c_{d'} $
  and
  $d'' \cdot a - c_{d''} $
  end up in the same subinterval/pigeonhole.
  But this means that
  $c = c_{d'} - c_{d''}$
  and
  $d = d' - d''$
  are integers such that
  $\abs{d \cdot a - c} \leq \frac{1}{n}$.

  We have now found integers $c$ and~$d$ such that
  $\abs{d \cdot a - c} \leq \frac{1}{n}$.
  Dividing this inequality by~$d$, we get that
  $\Abs{a - \frac{c}{d}} \leq \frac{1}{dn}$
  as desired.
  (Just in case we worry whether dividing like this is in order, we
  can see that the   inequality with absolute sign is equivalent to 
  the two inequalities
  $ 
  -\frac{1}{n}
  \leq
  d \cdot a - c
  \leq 
  \frac{1}{n}$,
  and dividing by the positive integer~$d$ yields
  $ 
  -\frac{1}{dn}
  \leq
  a - \frac{c}{d}
  \leq 
  \frac{1}{n}$,
  which is the same inequality with absolute sign as above.)
\end{solution}

\end{problem}



\begin{problem}%
  \label{problem:relations}%
  (70 p)
%%% 20 p for first two subproblems, 30 p for 3rd problem
  In this problem we focus on relations.
%     In particular,
  Suppose that
  $A = \set{e_0, e_1, \ldots, e_5} $   
  is a set of $6$~elements and consider the relation~$R$
  on~$A$
  represented by the matrix
  \begin{equation*}
    M_R = 
    \begin{pmatrix}
      0 & 0 & 1 & 0 & 0 & 0 \\
      0 & 0 & 0 & 1 & 0 & 0 \\
      0 & 0 & 0 & 0 & 1 & 0 \\
      0 & 0 & 0 & 0 & 0 & 1 \\
      1 & 0 & 0 & 0 & 0 & 0 \\
      0 & 1 & 0 & 0 & 0 & 0 
    \end{pmatrix}
  \end{equation*}
  (where element $e_i$ corresponds to row and column~$i+1$).

  \begin{subproblem}
%       (20 p)
    Let us write~$S$ to denote the symmetric closure of the relation~$R$.
    What is the matrix representation of~$S$?
    Can you explain in words what the relation~$S$ is
    by describing how it can be interpreted?
  \end{subproblem}

\begin{solution}
  Let us identify the elements $e_0, e_1, \ldots, e_5$
  with the numbers $0, 1, \ldots, 5$
  for the purposes of this discussion.

%     Firstly, by 
  By
  studying the matrix~$M_R$ we can conclude that
  $a$ and~$b$ are related by~$R$---which we have learned
  to denote by $a R b$, or $(a,b) \in R$---precisely when
  $
  b \equiv a + 2 \pmod{6}
  $.
  Taking the symmetric closure means that 
  $a$ and~$b$ can switch places,
  so we get that
  $(a,b) \in S$
  if
  $b \equiv a \pm 2 \pmod{6}$,
  and
%     which is expressed by the matrix
  \begin{equation*}
    M_S = 
    \begin{pmatrix}
      0 & 0 & 1 & 0 & 1 & 0 \\
      0 & 0 & 0 & 1 & 0 & 1 \\
      1 & 0 & 0 & 0 & 1 & 0 \\
      0 & 1 & 0 & 0 & 0 & 1 \\
      1 & 0 & 1 & 0 & 0 & 0 \\
      0 & 1 & 0 & 1 & 0 & 0 
    \end{pmatrix}
  \end{equation*}
  is the matrix representation of this relation.
  
  We can also compute directly on the matrix and obtain
  $M_S$
  as the coordinate-wise Boolean or of
  $M_R$
  and~$(M_R)^{\top}$.
%     and~$(M_R)^{T}$.
\end{solution}

  \begin{subproblem}%
    \label{problem:trans-clos}%
%       (20 p)
   Now let~$T$ be the transitive closure of the relation~$S$.
    What is the matrix representation of~$T$?
    Can you explain in words what the relation~$T$ is
    by describing how it can be interpreted?
  \end{subproblem}
  
\begin{solution}
  As just noted, we have
  $(a,b) \in S$  
  if
  $
  b \equiv a \pm 2 \pmod{6}
  $.
  Now, if for $a, b, c$ it holds that
  $
  b \equiv a \pm 2 \pmod{6}
  $
  and
  $
  c \equiv b \pm 2 \pmod{6}
  $,
  then
  $c$ and~$a$ differ
  by $0$, $2$, or~$4$
  $(\bmod\ 6)$.
  Or, in other words,
  $a$~and~$c$ are both odd or both even.
  This shows that
  $S^2$ is the relation such that 
  $(a,b) \in S^2$ 
  if $a$ and~$b$ are both odd or both even.
  If 
  $(a,b) \in S^2$ 
  and
  $(b,c) \in S^2$, then again we can conclude that
  $a$~and~$c$ are both odd or both even, i.e., 
  that $(a,c) \in S^2$
  Hence, $T = S^2 = S^3$ is closed under transitivity, 
  and so is the transitive closure of~$S$.

  Since we have concluded that
  $
  (a, b) \in T
  $
  if $a$ and~$b$ are both odd or both even,
  we see that this relation can be represented as
  \begin{equation*}
    M_T = 
    \begin{pmatrix}
      1 & 0 & 1 & 0 & 1 & 0 \\
      0 & 1 & 0 & 1 & 0 & 1 \\
      1 & 0 & 1 & 0 & 1 & 0 \\
      0 & 1 & 0 & 1 & 0 & 1 \\
      1 & 0 & 1 & 0 & 1 & 0 \\
      0 & 1 & 0 & 1 & 0 & 1 
    \end{pmatrix}
  \end{equation*}
  in matrix form.
    
\end{solution}


  \begin{subproblem}
%       (30 p)
    Suppose that we instead let~$T'$ be the transitive closure of the
    relation~$R$, and then let $S'$ be the symmetric closure of~$T'$.
    Are $S'$ and $T$ the same relation? 
    If they are not the same, show some way in which they differ.
    If they are the same, is it true 
    that $S'$ and~$T$ constructed in this way from some relation~$R$
    on    a set~$A$ will always be the same?
    Please make sure to motivate your answers clearly.
  \end{subproblem}

\begin{solution}
  Applying the same kind of reasoning as in
  Problem~\ref{problem:trans-clos},
  we get that if
  $
  b \equiv a + 2 \pmod{6}
  $
  and
  $
  c \equiv b + 2 \pmod{6}
  $,
  then
  $
  c \equiv a \pm 2 \pmod{6}
  $,
  and one more application of transitivity gives us the relation~$T$.
  Since we already argued that $T$ is closed under transitivity, we
  have
  $T' = T$.
  This relation is also symmetric, so taking the symmetric closure
  does not change it. Hence,
  we indeed have
  $S' = T' = T$.

  This does not hold in general, however.
  Let 
  $A = \set{1,2}$
  and let $R$ be the relation containing only the pair
  $(1,2)$.
  Then taking the transitive closure does not affect anything,
  so the relation stays the same, 
  and taking the symmetric closure of the transitive closure yields the relation 
  $
  \set{
    (1,2), (2,1)
  }
  $
  with  two pairs.
  However, if we take the symmetric closure first, so that we get
  $
  \set{
    (1,2), (2,1)
  }
  $,
  then the transitive closure will also relate
  $1$ and~$2$ with themselves, so that the final relation 
  after symmetric closure followed by transitive closure 
  is
  $
  \set{
    (1,1), (1,2)
    (2,1), (2,2)
  }
  $.
  Expressed in in slightly more fancy language, we have now shown that
  the two operations of symmetric closure and transitive closure do
  not commute.
\end{solution}

\end{problem}



%%% Macros for logical connectives
\newcommand{\distinct}[2]{\mathsf{setsize}\left(#2, #1\right)}
\providecommand{\Land}{\bigwedge}
\providecommand{\limpl}{\to}
%%% No, \sim for negation just looks insane...
\providecommand{\mylnot}{\sim}
\renewcommand{\mylnot}{\lnot}
%%%

\begin{problem}%
  (120 p)
  Recall that an undirected graph $G = (V,E)$ consists of a set of
  vertices $V$ connected by edges $E$, where every edge is an
  unordered pair of vertices. If there is an edge $(u,v)$ between two
  vertices $u$ and $v$, then we say that $u$ and $v$ are the \emph{endpoints}
  of the edge, and the two vertices are said to be \emph{neighbours}.  
  We say that a  sequence of edges
  $
  (v_1, v_2), \,
  (v_2, v_3), \,
  (v_3, v_4), \,
  \ldots, \,
  (v_{k-1}, v_k), 
  $
  in~$E$
  is a \emph{path} from $v_1$ to~$v_k$.

  In this problem, we wish to express properties of graphs in both
  natural language and predicate logic, and to translate between
  the two forms. 
  We do this as follows:
  \begin{itemize}
  \item 
    The universe is  the set of vertices $V$ of $G$.
  \item
    The binary predicate $E(u,v)$ holds if and only if there is an
    edge between $u$ and $v$ in $G$. 
  \item
    The unary predicate $S(v)$ is used to identify a subset of vertices
    $S = \{v \mid v \in V, \, S(v) \text{ is true}\}$
    for which some property might or might not hold.
  \end{itemize}
  For example, we can write
  the natural language statement 
  ``$S$ is a set containing exactly $k$~distinct elements''
  as a 
%   predicate
  formula
  \begin{equation}
    \begin{split}
      \distinct{k}{S} 
    \ := \qquad
  %   \eqdef{} 
    &
    \exists u_1 \cdots \exists u_k 
    \left( \Land_{i=1}^{k-1}\Land_{j=i+1}^{k} \l(u_i \neq u_j) 
     \land \Land_{i=1}^{k} S(u_i) 
    \right)\\
    \land \ &
    \lnot \exists u_1 \cdots \exists u_{k+1} 
    \left( \Land_{i=1}^{k}\Land_{j=i+1}^{k+1} \l(u_i \neq u_j) 
     \land \Land_{i=1}^{k} S(u_i) 
    \right)
    \end{split}
%       \eqcomma
  \end{equation}
  where $\Land_{i=1}^k \phi_i$ is shorthand for $\phi_1 \land \phi_2
  \land \cdots \land \phi_k$,
  and where we use standard notation
  $\lnot$
  for logical negation.
%%%
  In natural language, this formula can be read as: 
  ``There exist $k$ elements $u_1$ to $u_k$ such that 
  (i)~for every  pair $(u_i,u_j)$,
  $i \neq j$,
  the elements themselves are also distinct ($u_i\neq u_j$); and
  (ii)~all the $u_i$ for $i=1,\ldots,k$ are members of $S$,
  but no such set of 
  $k+1$ elements $u_1$ to $u_{k+1}$ exists.''


  Below you find 
  six graph properties defined in natural language
  and
  six graph properties written as predicate logic formulas.
  Most of the natural language definitions have equivalent predicate
  logic formulas, but not all.

  Your task is to translate each predicate logic formula
  (a), \ldots, (f) into a natural language description, and argue
  which---if any---of the natural language definitions (1), \ldots, (6)
  it matches.

\bigskip

\noindent
\textbf{Natural Language Definitions:}
\begin{enumerate}[(1)]
    \item 
      A \emph{dominating set} of size~$k$ for a graph $G=(V,E)$ is a set $S$ of $k$ 
        distinct vertices such that every vertex $v$ in the graph either is in $S$ or is a neighbour of a vertex in  $S$.
      \item
        A \emph{clique} $S$ of size~$k$ in a graph $G=(V,E)$ is a set $S$ of $k$ 
        distinct vertices such that all vertices in $S$ are neighbours with each other. 
    \item  %%% NOT MATCHED
      A \emph{disconnected vertex set} of size~$k$ in a graph $G=(V,E)$ is a
      set $S$ of $k$ distinct vertices such that there are no edges from any
      $u \in S$ to any $v \in V \setminus S$.
      [Here $V \setminus S$ denotes set subtraction, so that
      $V \setminus S = \{ v \mid v \in V \text{ and } v \notin S \}$.]
        
    \item 
      A \emph{vertex cover} of size~$k$ of a graph $G=(V,E)$ is a set $S$ of $k$ 
      distinct vertices such that for every edge $(u,v) \in E$ it holds that at least one of the endpoints is in $S$.
    
    \item 
      A graph $G = (V,E)$ is \emph{bipartite},
      with one of the parts in the bipartition having size~$k$,
      if there a set $S$ of $k$ 
      distinct vertices such that all edges in the graph go between $S$ and $V \setminus S$.

    \item %%% NOT MATCHED
      A \emph{connected component} $S$ of size~$k$ in a graph $G=(V,E)$ is a set $S$ of $k$ 
        distinct vertices such that for every pair of distinct vertices $u$ and $v$ in $S$ there is a path from $u$ to~$v$
        consisting of vertices in $S$.        
\end{enumerate}


\bigskip

\noindent
\textbf{Predicate Logic Formulas:}
\begin{enumerate}[(a)]
\item %% Vertex cover --- match (4)-(a)
      $
\distinct{k}{S} 
    \, \land \,
    \forall v \forall w \bigl( E(v,w) \limpl ( S(v) \lor S(w)  ) \bigr)
    $

    
    \item %% Dominating set --- match (1)-(b)
    $
\distinct{k}{S} 
    \, \land \,
    \forall v \bigr( S(v) \lor \exists w (S(w) \land E(v,w)) \bigl)
    $
    
    \item %% Unmatched! Intended to maybe look a bit similar to connectivity, but the formula is not right
    %% S is a vertex set of size k such that any two distinct vertices in S have a common neighbour in the graph
    $
\distinct{k}{S} 
    \, \land \,
    \forall u \forall w \bigl( ( u \neq w \land  S(u) \land S(w) )
    \limpl
    \exists v (E(u, v) \land E(v,w ) \bigr)
    $

    \item %% Bipartiteness --- match (5)-(d)
    $
\distinct{k}{S} 
    \, \land \,
    \forall v \forall w \bigl( E(v,w) \limpl 
    (
    (S(v) \land \mylnot S(w)) 
    \lor 
    (\mylnot S(v) \land S(w)) 
    )
    \bigr)
    $

    %% Unmatched! NOT disconnected set, but somewhat close except for obvious mistake with wrongly placed negation in the encoding
    %% S is a vertex set of size k such that every vertex v in S has an edge to some vertex w outside of S 
  \item
    $
    \distinct{k}{S}
    \, \land \,
    \forall v \bigl( S(v) \limpl \exists w (\mylnot S(w) \land E(v,w)) \bigr)
    $
    
    \item %% Clique --- match (2)-(f)
    $
\distinct{k}{S} 
    \, \land \,
    \forall v \forall w \bigl( ( v \neq w \land  S(v) \land S(w)) \limpl E(v, w) \bigr)
    $

\end{enumerate}

\medskip
\noindent
For every predicate logic formula that matches 
a natural language description, explain clearly why there is a match.
For formulas that do not match a description, write a natural language
definition along the lines of (1), \ldots, (6)
that describes the property that the formula encodes.




\begin{solution}
Formula 
(a) matches the description (4) of vertex cover. It says that if there is an edge between two vertices $v$ and $w$,
then at least one of these vertices is in~$S$. This is the definition of $S$ being a vertex cover.  

Formula 
(b) matches the description (1)  of dominating set. It says that for every vertex~$v$ it holds that either $v$ is in the set $S$
or there is some vertex~$w$ in $S$ that is connected to $v$ via an edge. This precisely matches the given description of what a dominating set is.

Formula 
(c) is unmatched. It encodes that
$S$ is a vertex set of size $k$ such that any two distinct vertices in $S$ have a common neighbour in the graph.

Formula 
(d) matches the description (5) of bipartiteness. It says that if there is an edge between $v$ and $w$,
then either $v$ is in $S$ and $w$ is not, or the other way round. That is, exactly one of the vertices is any edge is in~$S$,
which shows that $(S, V\setminus S)$ is indeed a bipartition of the graph.

Formula 
(e) is unmatched. It encodes that
$S$ is a vertex set of size $k$ such that every vertex~$v$ in $S$ has an edge to some vertex $w$ outside of $S$. 

Formula (f) matches the description (2) of clique.  It says that if
two distinct vertices $v$ and~$w$ are both in $S$, then there has to
be an edge between them.  This is exactly the condition for vertices
in a clique.
\end{solution}

\end{problem}





\end{document}


